---
output:
  html_document: default
  html_notebook: default
---

EDA of Loan Data by Amanullah Ashraf
========================================================


  ```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
  
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(alr3)
library(reshape2)
library(dplyr)
library(GGally)
library(RColorBrewer)
library(FField)
library(ggrepel)

```


```{r echo=FALSE, Load_the_Data}

# Load the Data
ld <-read.csv('prosperLoanData.csv')

```

This data set is about the people bought loan. How much is their loan, their credit grade,
occupation, income range, loan amout, stated monthly income, employment status etc.


# Univariate Plots Section

## Summary of Loan Status 

```{r echo=FALSE}


# Summary of different columns

summary(ld$LoanStatus)

```

## Summary of Loan Original Amount
```{r  echo=FALSE}

summary(ld$LoanOriginalAmount)

```

## Mean, Median
```{r  echo=FALSE}

mean(ld$LoanOriginalAmount) # Mean of LoanOriginalAmount


median(ld$LoanOriginalAmount) # Median of LoanOriginalAmount
```


## Range of Loan Original Amount

```{r  echo=FALSE}

range(ld$LoanOriginalAmount) # Range of values in LoanOriginalAmount column

```

## Range of Stated Monthly Income
```{r  echo=FALSE}

range(ld$StatedMonthlyIncome) # Range of values in StatedMonthlyIncome column

```

## Range of Term Column

```{r  echo=FALSE}

range(ld$Term) # Range of values in EmploymentStatusDuration column

```

# Plotting histogram of different columns

```{r  echo=FALSE}

qplot(x=LoanStatus, data=ld) + 
  #scale_x_discrete(labels = abbreviate) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1))

```

Above graph shows current loan status of the borrower

```{r  echo=FALSE}

qplot(x=Occupation, data=ld) +
  #scale_x_discrete(labels = abbreviate) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1))

```

Above graph shows different occupation of the borrowers


```{r  echo=FALSE}

qplot(x=EmploymentStatus, data=ld) 

```

Above graph shows current employment status of the borrower


```{r  echo=FALSE}

qplot(x=EmploymentStatusDuration, data=ld)

```

Above graph shows for how long borrower has been employed

```{r  echo=FALSE}

qplot(x=IncomeRange, data=ld) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1))

```

Above graph shows income range of the borrowers

```{r  echo=FALSE}

qplot(x=CreditGrade, data=ld) 

```

Above graph shows credit grade of the borrower

```{r  echo=FALSE}

qplot(x=LoanOriginalAmount, data=ld) 

```

Above graph shows how much loan of original amount taken

```{r  echo=FALSE}

qplot(x=BorrowerState, data=ld)  + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1))

```

Above plot shows which state has highest borrowing

```{r  echo=FALSE}

ggplot(aes(x=Term), data=ld) +
  geom_histogram(binwidth = 25)

```

Above plot shows duration of the loan

```{r  echo=FALSE}

qplot(x=IsBorrowerHomeowner, data=ld) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1))

```

Above graph shows the purpose for taking a loan

```{r  echo=FALSE}

qplot(x=LoanOriginationQuarter, data=ld)+ 
  scale_x_discrete(labels = abbreviate) + 
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1))

```

Above graph shows which quarter of the year loan amount was taken

# Univariate Analysis



### What is the structure of your dataset?

Loan data set is compsed of 113937 obersvation and 81 variables. In the data set there 

are values which are numceric and other are categorical. It contain information about

peolple who took loan and their current ststus. It contain information about their occupation

thier income, state, credit rating, loan status, employment staus etc.  

### What is/are the main feature(s) of interest in your dataset?

Borrower Occupation, thier income range and how much loan original amount they got and how

many have paid them back. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Other columns like borrower credit rating, borrower state, stated monthly income etc are

helpful as well in investigation. 

### Did you create any new variables from existing variables in the dataset?

No new variable is created as data set already contain quite large number of variables. 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Colums which I am using is quite clean. I didnt feel any need to clean it.  

# Bivariate Plots Section

## Employment Status wise loan status

```{r echo=FALSE, Bivariate_Plots}

by(ld$LoanStatus, ld$EmploymentStatus, summary) 

```

## This shows LoanStatus wise LoanOriginalAmount where loan is still due

```{r echo=FALSE}

by(ld$LoanOriginalAmount, ld$LoanStatus=='Current', summary) 

```


# Checking Relationship between different vriables

## Checking Relationship between StatedMonthlyIncome & LoanOriginalAmount

```{r echo=FALSE}

cor.test(ld$StatedMonthlyIncome, ld$LoanOriginalAmount, method='pearson')

```

## Checking Relationship between Term & LoanOriginalAmount

```{r echo=FALSE}

cor.test(ld$Term, ld$LoanOriginalAmount, method='pearson')

```

## Checking Relationship between EmploymentStatusDuration & LoanOriginalAmount

```{r echo=FALSE}

cor.test(ld$EmploymentStatusDuration, ld$LoanOriginalAmount, method='pearson')

```


## Checking Relationship between StatedMonthlyIncome & EmploymentStatusDuration

```{r echo=FALSE}

cor.test(ld$StatedMonthlyIncome, ld$EmploymentStatusDuration, method='pearson')

```


# Histogram of 2 variables


```{r echo=FALSE}

qplot(  x = ld$LoanOriginalAmount,
        data = ld, 
        xlab ='LoanOriginalAmount(USD)',
        ylab='count',
        binwidth = 2500,
        geom='freqpoly',
        color=LoanStatus) +
  scale_x_continuous(limits = c(0, 35000), breaks = seq(0, 35000, 10000))

```

The above graph showes borrower loan status and their loan original amount in order 
to check the current status, defaulters and who had alreay paid the amount.

```{r echo=FALSE}


  qplot(x=Term, y=LoanOriginalAmount,
      data=ld, geom='boxplot') +
  geom_point(alpha=1/20 , position=position_jitter(h=0), color='orange')


```

This box plot is showing 5 number summary of Term column against Loan original amount.



```{r echo=FALSE}

ggplot(aes(x=StatedMonthlyIncome, y=LoanOriginalAmount), data=ld) +
  geom_point(fill=I("#F79420"), color=I("black"), shape=21) +
  stat_smooth(method="lm") +
  scale_x_continuous(lim = c(0, quantile(ld$StatedMonthlyIncome, 0.99)) ) +
  scale_y_continuous(lim = c(0, quantile(ld$LoanOriginalAmount, 0.99)) ) +
  ggtitle("LoanData: StatedMonthlyIncome vs. LoanOriginalAmount(USD)")

```

This scatter plot shows borrower original Loan amount and their montly income, 
inorder to find out the loan trend among the monthly income borrowers. As the
monthly rises, loan amount is also rising with it. Means, more monthly income
you have, more loan amount you can get.


```{r echo=FALSE}

# Grouping by BorrowerState

ld_by_State <- group_by(ld, BorrowerState)
ld_mp_by_State <- summarise(ld_by_State, mean = mean(LoanOriginalAmount))
ld_mp_by_State

# Grouping by EmploymentStatus

ld_by_EmpStat <- group_by(ld, EmploymentStatus)
ld_mp_by_EmpStat <- summarise(ld_by_EmpStat, mean = mean(LoanOriginalAmount))
ld_mp_by_EmpStat

# Assign grouping to variables

plot1 <-  ggplot(aes(x=mean), data=ld_mp_by_State) + 
  geom_histogram(binwidth=200) +
  ggtitle('Mean loan by State')

plot2 <- ggplot(aes(x=mean), data=ld_mp_by_EmpStat) + 
          geom_histogram(binwidth=200) +
          ggtitle('Mean loan by EmpStatus')

# Plotting side by side in order to view

grid.arrange(plot1, plot2, ncol=2)

```

Plotting side by side in order to view Mean of loan by state and by Employment status.

```{r echo=FALSE}

ggplot(aes(x = LoanOriginalAmount), data = subset(ld, !is.na(LoanStatus))) +
  geom_freqpoly(aes(color = LoanStatus)) +
  scale_x_log10()

```

```{r echo=FALSE}

qplot(x = LoanOriginalAmount, data = ld) + facet_wrap(~LoanStatus) +
  scale_x_log10()

```

Histogram of loan amount according to loan status

```{r echo=FALSE}

qplot(x=LoanStatus, y=LoanOriginalAmount,
      data=ld, geom='boxplot') + 
  geom_point(alpha=1/20 , position=position_jitter(h=0), color='orange') +
    theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1))

```

This box plot is showing 5 number summary of loan status colum against loan original amount.

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I tried to find out the different combination of coulms with the Loan Status. For example,

Loan Status with loan original amount, loan status with with occupation, loan status with 

employment status etc. Drew different kinds of plots.

### Did you observe any interesting relationships between the other features \
 (not the main feature(s) of interest)?

Yes, the relation between EmploymentStatusDuration and LoanOriginalAmount and other one

between StatedMonthlyIncome and EmploymentStatusDuration.

### What was the strongest relationship you found?

The strogest relationship exist between Term and LoanOriginalAmount, 

i.e 0.3389275

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

# Creating a new data frame

ld.fc_by_Loanamt_Occ <- ld %>%
  filter(!is.na(EmploymentStatus)) %>%
  group_by(EmploymentStatus, Occupation, LoanOriginalAmount) %>%
  summarise(mean_loanamt=mean(LoanOriginalAmount),
            median_loanamt=median(LoanOriginalAmount),
            n=n()) %>%
  ungroup() %>%
  arrange(LoanOriginalAmount)

# Drwaing line grpah between LoanOriginalAmount, Occupation, EmploymentStatus. 

ggplot(aes(x = Occupation , y = LoanOriginalAmount),
       data = subset(ld.fc_by_Loanamt_Occ, !is.na(EmploymentStatus))) +
  geom_line(aes(color = EmploymentStatus)) +
  theme(text = element_text(size=9),
        axis.text.x = element_text(angle=90, hjust=1))

```

Drwaing line grpah between LoanOriginalAmount, Occupation, EmploymentStatus. This will 
show the employment status of the borrower, thier occupation and thier loan amount.

```{r echo=FALSE}

# Creating a new data frame

ld.fc_by_Loanamt_Occ.wide <- dcast(ld.fc_by_Loanamt_Occ,
                                   LoanOriginalAmount ~ EmploymentStatus,
                                   value.var = 'median_loanamt')
#head(ld.fc_by_Loanamt_Occ.wide)



ggplot(aes(x = LoanOriginalAmount, y = Employed),
       data = ld.fc_by_Loanamt_Occ.wide) +
  geom_line()
  # geom_hline(yintercept = 1, alpha=0.3, linetype=2)

```

This line graph is showing the employed borrower loan amount. Most of the 
people on average took below $10,000 loan amount, as there are more black
lines in the range of $0 to $10,000



# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Yes, the relationship between loan amount, employment status and occupation. 

### Were there any interesting or surprising interactions between features?

Yes, borrower who are retired were also given loan amount.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary



### Plot One


```{r echo=FALSE, Plot_One}

qplot(x=IncomeRange, y=LoanOriginalAmount, data=ld, geom='boxplot') +
  ggtitle("IncomeRange vs. LoanOriginalAmount(USD)") +
  theme(text = element_text(size=10),
        axis.text.x = element_text(angle=90, hjust=1)) +
  geom_point(alpha=1/20 , position=position_jitter(h=0), color='green')
  

```

### Description One

Borrower of different income range are shown on the X-axis and their corresponding loan amount are on the Y-axis. 
From the above graph we can say that there is a kind of symetry of the loan amount taken. Except $100,000+ income 
range group, almost all other group took, $25000 loan amount. The highest loan amount taken by $100,000 income range
group i.e $35,000. They are good traget market segmentation for the lender. Lender can offer them different products
inorder to retain them.

### Plot Two

```{r echo=FALSE, Plot_Two}

ggplot(aes(x = StatedMonthlyIncome, y = LoanOriginalAmount), data = ld) + 
  geom_point(alpha=0.09 , position=position_jitter(h=0), color='orange') + 
  geom_line(stat='summary', fun.y=mean,  color='green') +
 # geom_line(stat='summary', fun.y=quantile, fun.args = list(probs = .1), linetype=2,   color='blue') +
 # geom_line(stat='summary', fun.y=quantile, fun.args = list(probs = .5), linetype=2, color='black') +
 # geom_line(stat='summary', fun.y=quantile, fun.args = list(probs = .9), linetype=2, color='red') +
  scale_x_continuous(limits = c(0, 200000), breaks = seq(0, 200000, 25000)) + 
  ggtitle("LoanData: StatedMonthlyIncome vs. LoanOriginalAmount(USD)")

```

### Description Two

From the above graph we can say that most of the loan are taken by those whose monthly income is from $0 to $25,000. 
This group almost took $35,000 loan amount. The second major chunck on X-axis is from From 25,000 to 50,000. Except few 
exceptions, the second group on average took $25000 loan amount. One major thing to note from the above graph is, those 
whose monthly income ranging from $0 to $25000 took most of the loan. One reson might be that they have low monthly 
income, so inorder to meet their expenses they need to take loan. Lender can target this group in their markting stratigies.

### Plot Three

```{r echo=FALSE, Plot_Three}

ld.fc_by_Loanamt_Occ.wide <- dcast(ld.fc_by_Loanamt_Occ,
                                   LoanOriginalAmount ~ EmploymentStatus,
                                   value.var = 'median_loanamt')
head(ld.fc_by_Loanamt_Occ.wide)


Plot3 <- ggplot(aes(x = LoanOriginalAmount, y = Employed),
         data = ld.fc_by_Loanamt_Occ.wide) +
         geom_line()+
         geom_hline(yintercept = 1, alpha=0.3, linetype=2) + 
         ggtitle("LoanOriginalAmount vs. Employed")

Plot4 <- ggplot(aes(x = LoanOriginalAmount, y = Retired),
         data = ld.fc_by_Loanamt_Occ.wide) +
         geom_line()+
         geom_hline(yintercept = 1, alpha=0.3, linetype=2) + 
         ggtitle("LoanOriginalAmount vs. Retired")

grid.arrange(Plot3, Plot4, ncol=2)  

```

### Description Three

From the above graph we can see that loan amounts were given to both Employed and Retired persons. We can see that loan
amount to Retired people were $10000 majority of the times, but people were given in $10,000 to $25000 range as well.
employed people are all over the placeses on the graph. This shows that employed people are availaing more loan than
retired people. Lender can come up with some good product for employed segment of loan borrower.


# Reflection

From the graphs drwan above we can observe the following information:
According to loan status, majority of the people fall under "Currently" section, means currently majority of people have taken loan 
who are employed under the income range of $25000 to $49,999. Majority of the people took upto $25000 loan amount. State wise, people 
from California took most of the loan. Almost half of the population of the borrower are home owners. Occupation wise highest among 
them who took loan is Nurse(RN). 
So we can hypothesise by saying that, currently people are under debt, mostly from california who are employed under the Nurse(RN) 
occupation with the monthly income range of $25000 > to $49,999. We can further test our hypothesis by doing regression analysis. 

Plot 1 shows income range of the people and their borrwed amount. One of the income range group took $35000 loan.
They are good traget market for lender. Lender can introduce new products to them, offer discount, offer prizes etc.  
From the above plot 2 we can say that most of the loan are taken by those whose monthly income is from $0 to $25,000.
Low imcome are majority who avail the loan opportunity. 
Plot 3 shows that Employed people are availaing more loan than retired people. Why is that? Are they unable to meet 
their expenses from their job as they are employed or inflation is high which in turn forced them to take loan? 
From so many columns it is difficiult to a select few colums for analysis. Good thing about Data set is, it is almost 
clean, while analyzing. It is a surprises to know that people who are retired also got loan amount. 
